---
title: "AR(1) power study and algorithm checks"
author: "svpChange2"
date: "2026-09-15"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
project_root <- normalizePath(
  file.path(dirname(knitr::current_input(dir = TRUE)), "..", "..")
)
knitr::opts_knit$set(root.dir = project_root)
setwd(project_root)
library(svpChange2)
source(file.path("simulations", "power_common.R"))
source(file.path("simulations", "power_ar1", "run_power.R"))
```

# Study model

The study uses

\[
Y_t = \mu_t + E_t, \qquad
E_t = \rho E_{t-1} + \eta_t,
\qquad \eta_t \sim N(0,\sigma^2).
\]

It fixes `rho = 0.8` and requires marginal AR(1) noise variance one. Hence

\[
\sigma^2 = 1 - \rho^2 = 0.36.
\]

`ts_generator()` receives `sd_noise = sqrt(sigma2)` and initializes the first
residual from the stationary distribution. The simulation therefore has
marginal variance one, not innovation variance one.

```{r variance-check}
set.seed(20260915)
sample_variances <- replicate(
  200L,
  var(simulate_ar1_noise(600L, rho = 0.8))
)
mean(sample_variances)
range(sample_variances)
stopifnot(abs(mean(sample_variances) - 1) < 0.05)
```

# Signal scenarios and study size

The four scenarios are:

* `none`: one constant segment;
* `up`: eight equally long segments with levels
  `0, jump, 2 * jump, ..., 7 * jump`;
* `updown`: eight equally long segments alternating between `0` and `jump`;
* `rand1`: eight reproducibly random segment lengths and signed levels.

The corrected grid is `seq(1, 5, 0.2)`. It contains 21 levels, giving 8,400
series and 42,000 fits:

```{r design-check}
jump_grid <- seq(1, 5, 0.2)
length(jump_grid)
stopifnot(identical(length(jump_grid), 21L))
stopifnot(isTRUE(all.equal(jump_grid, seq(1, 5, 0.2))))
```

The larger upper limit compensates for the lower effective information about
a mean change caused by serial dependence. A corrected-scale pilot reached
F1 values near 0.96 around jump 4 and near 1 around jump 5 in the difficult
`up` and `updown` scenarios.

# Algorithms in the power study

## Inflated Gaussian PELT

`PELT inflated` applies Gaussian mean PELT to the original observations. It
does not use the AR(1) likelihood. Dependence is handled only by increasing
the penalty to

\[
3\log(n)\frac{1+\rho}{1-\rho}.
\]

For `rho = 0.8`, this is `27 * log(n)`, which is deliberately conservative.

## Approximate AR(1) PELT

The approximate AR(1) method transforms the observations into

\[
Z_t = Y_{t+1} - \rho Y_t,
\qquad t=1,\ldots,n-1.
\]

It then calls the existing package function `PELT()` on `Z`:

```{r transformed-pelt}
set.seed(12)
n <- 180L
rho <- 0.7
y <- ts_generator(
  chpts = c(60L, 120L, n),
  parameters = c(0, 4, -2),
  sd_noise = sqrt(1 - rho^2),
  rho = rho,
  type = "gaussAR1"
)
z <- y[-1L] - rho * y[-n]
penalty <- 2.2 * log(n)
pelt_z <- svpChange2::PELT(z, penalty)
op_z <- svpChange2::OP(z, penalty)
stopifnot(identical(pelt_z$changepoints, op_z$changepoints))
stopifnot(isTRUE(all.equal(pelt_z$costQ, op_z$costQ, tolerance = 1e-10)))
pelt_z$changepoints
```

Thus this is genuinely PELT on transformed data. It is called *approximate*
because the transformation does not retain the complete exact AR(1)
likelihood treatment at every possible segment boundary. PELT supplies both
the optimization and the pruning; no second PELT implementation is used.

An endpoint `j` in `z` corresponds to original endpoint `j + 1`. The
transformed terminal endpoint is removed and the original endpoint `n` is
appended:

```{r transformed-boundary-map}
internal_z <- head(pelt_z$changepoints, -1L)
mapped_y <- normalise_boundaries(internal_z + 1L, n)
mapped_y
```

The existing PELT diagnostics expose the candidate workload:

```{r pelt-diagnostics}
sum(pelt_z$nb)
length(pelt_z$lastIndexSet)
```

## DeCAFS AR(1)

`DeCAFS AR1` uses the DeCAFS state-space algorithm with:

```r
modelParam = list(
  sdEta = 0,
  sdNu = sqrt(1 - rho^2),
  phi = rho
)
```

`sdEta = 0` disables the random-walk component. Its penalty is calibrated
independently under the AR(1) null. The selected multiplier is
`4.75 * log(n)`, the smallest tested value achieving the 95% no-change target.

## SVP AR1Focus

`SVP AR1Focus` uses the incremental AR1Focus validity statistic with:

```r
gamma = 3.75 * log(n)
subtests = "right"
```

`SVP AR1Focus multiscale` uses the same statistic with:

```r
  gamma = 4.5 * log(n)
  subtests = "both"
```

The two SVP methods differ in their threshold and validity-pruning mode; they
are not two different AR(1) likelihood models. SVP first minimizes the number
of valid segments and then uses within-segment SSE as its tie-breaker.

The exact fixed-variance and profiled AR(1) tests are available as
`test = "AR1"` and `test = "AR1Profile"`, but are not part of this five-method
power grid.

# One-series comparison

All five methods receive the same observations:

```{r one-series}
set.seed(20260915)
mu <- generate_power_signal(600L, "up", 4)
y <- mu + simulate_ar1_noise(600L, rho = 0.8)
fits <- fit_ar1_methods(y, rho = 0.8)
lapply(fits, identity)
```

# Reproducing the full power study

The full study is opt-in and can be run from the package root:

```{r full-study, eval=FALSE}
Sys.setenv(SVP_RUN_SIMULATIONS = "true")
source("simulations/power_ar1/run_power.R")
run_and_save_ar1(workers = 1L)
```

The generated files contain the corrected calibration, power results, plots,
and direct changepoint outputs. All methods use the same seed for each design
row, so comparisons are paired on identical simulated data.

# Conclusion

The corrected study has marginal AR(1) variance one and innovation variance
`1 - rho^2`. The approximate AR(1) comparator is the existing package PELT
applied to transformed data, AR1Focus is the incremental SVP validity method,
and DeCAFS is a separate state-space segmentation algorithm.
