---
title: "The three `subtests` modes of `SVP()`"
author: ""
date: "14 September 2026"
output:
  pdf_document:
    latex_engine: pdflatex
fontsize: 10pt
geometry: [a4paper, margin=2cm]
header-includes:
  - \usepackage{amsmath,amssymb,booktabs,array,float}
  - \DeclareMathOperator*{\arglexmin}{arg\,lex\,min}
  - \DeclareMathOperator*{\lexmin}{lex\,min}
  - \setlength{\parindent}{0pt}
  - \setlength{\parskip}{3pt}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.width = 6.7, fig.height = 2.45,
  fig.pos = "H", out.width = "96%"
)
library(svpChange2)

project_root <- normalizePath(file.path("..", ".."))
knitr::opts_knit$set(root.dir = project_root)
modes <- c("none", "right", "both")
endpoint_text <- function(fit) paste(fit$changepoints, collapse = ", ")
time_root <- file.path(project_root, "simulations", "time_gaussian")
comparison_path <- file.path(
  time_root, "gaussian_time_shortcuts_comparison.csv"
)
summary_path <- file.path(time_root, "gaussian_time_shortcuts_summary.csv")
stopifnot(file.exists(comparison_path), file.exists(summary_path))
time_comparison <- read.csv(comparison_path, stringsAsFactors = FALSE)
time_summary <- read.csv(summary_path, stringsAsFactors = FALSE)
```

Version 0.2.0 accepts three values of `subtests`: `"none"`, `"right"`, and
`"both"`. The former `"left"` label was removed because, in the incremental
`SVP()` implementation, it executed exactly the same inclusive-left branch as
`"both"`. It is also rejected by `svp0()` so that the two public interfaces
have the same choices. This change does not concern the internal left-hand
geometric states used by FOCuS.

## Common recursion

Let $y_{s..t}=(y_{s+1},\ldots,y_t)$, where $s$ is a zero-based boundary and
$t$ is the one-based inclusive endpoint. Write

\[
\mathcal C(s,t)=\sum_{i=s+1}^{t}(y_i-\bar y_{s..t})^2
\quad\text{and}\quad
V_\gamma(s,t)=\boldsymbol 1\{y_{s..t}\text{ is valid}\}.
\]

Singletons are always valid. For longer segments, the compiled test must be
finite and strictly below `gamma`. Both public routines solve the
lexicographic problem

\begin{equation}
\label{pb:svp}
\widehat\tau\in
\arglexmin_{\substack{0=\tau_0<\cdots<\tau_K=n\\
V_\gamma(\tau_{j-1},\tau_j)=1,\ j=1,\ldots,K}}
\left(K,\sum_{j=1}^{K}\mathcal C(\tau_{j-1},\tau_j)\right).
\end{equation}

With $R_t=(K_t,Q_t)$ and $R_0=(0,0)$, the unpruned recursion is

\begin{equation}
\label{eq:recursion}
R_t=
\lexmin_{\substack{0\leq s<t\\V_\gamma(s,t)=1}}
\left\{R_s+\bigl(1,\mathcal C(s,t)\bigr)\right\}.
\end{equation}

Thus $K_t$ is minimized first; $Q_t$ breaks ties among partitions having the
same $K_t$. Candidate boundaries are compared in increasing order and the C++
cost comparison is strict. Consequently, an exact floating-point tie retains
the smallest visited boundary. Backtracking the predecessor column of `R`
from $n$ gives `changepoints`, including the final endpoint $n$.

Every active boundary $s$ in `SVP()` owns an incremental validity-test state,
the last observation fed to it, and its current validity. A deferred state is
caught up by feeding each missing observation once. Ordinary catch-up checks
validity only at the requested endpoint; the special right-pruning replay
checks every missed endpoint. Centered first and second prefix sums give
$\mathcal C(s,t)$ in $O(1)$ while avoiding translation-dependent numerical
cancellation.

## Pruning assumptions and three modes

Put $I_\gamma(s,t)=1-V_\gamma(s,t)$. The right-extension implication is

\begin{equation}
\label{eq:right}
I_\gamma(s,t)=1\quad\Longrightarrow\quad
I_\gamma(s,u)=1,\qquad u>t,
\end{equation}

and the left-extension implication is

\begin{equation}
\label{eq:left}
I_\gamma(s,t)=1\quad\Longrightarrow\quad
I_\gamma(r,t)=1,\qquad r<s.
\end{equation}

These are sufficient conditions for the corresponding permanent deletions to
preserve problem~\eqref{pb:svp}. They are properties of the complete validity
rule, not of the secondary SSE criterion. With an arbitrary user rule,
`"none"` is the reference unless the implications have been established.

Let $\mathcal A_{t-1}$ be the ordered candidate set at the beginning of
endpoint $t$, and let
$\mathcal F_t=\{s\in\mathcal A_{t-1}:V_\gamma(s,t)=0\}$.

| `subtests` | Permanent candidate action | Sufficient condition for equality with the unpruned recursion |
|:--|:--|:--|
| `"none"` | Keep every boundary. | None. |
| `"right"` | Delete each currently invalid boundary. | Equation \eqref{eq:right}. |
| `"both"` | Delete the largest invalid boundary and every older boundary. | Equations \eqref{eq:right} and \eqref{eq:left}. |

### `subtests = "none"`: segment-count buckets

Here $\mathcal A_t=\{0,\ldots,t\}$. The implementation groups existing
boundaries by their already computed segment count,

\[
\mathcal B_{t-1,k}=\{s<t:K_s=k\},
\qquad
k_t^\star=\min\{k:\exists s\in\mathcal B_{t-1,k},\ V_\gamma(s,t)=1\}.
\]

Equation~\eqref{eq:recursion} immediately gives

\[
K_t=k_t^\star+1,
\qquad
Q_t=\min_{\substack{s\in\mathcal B_{t-1,k_t^\star}\\
V_\gamma(s,t)=1}}
\{Q_s+\mathcal C(s,t)\}.
\]

The code therefore tests buckets in increasing $k$, stops after the first
bucket containing a valid segment, and compares every valid SSE in that bucket.
All larger-$k$ buckets are irrelevant to both coordinates. Skipped incremental
states remain stored and catch up if a later endpoint needs them. This is an
exact lexicographic shortcut and does not assume that $(K_t)$ is monotone. In
particular, if boundary zero is valid, only that test is needed and $K_t=1$.

### `subtests = "right"`: one-segment deferral and cost skipping

Outside its initial shortcut, right pruning applies

\[
\mathcal A_t=
\{s\in\mathcal A_{t-1}:V_\gamma(s,t)=1\}\cup\{t\}.
\]

While $V_\gamma(0,t)=1$, boundary zero gives the unbeatable value $K_t=1$.
Later tests are deferred, although their candidates remain physically stored.
When zero first fails, `replay_right_candidate()` reconstructs every deferred
history and deletes a candidate if it failed at any missed endpoint. This is
necessary: checking only its current endpoint would not reproduce right
pruning when validity can recover.

Thereafter every survivor is tested and retained. During the ascending scan,
once a provisional segment count $K_t^{\rm best}$ exists, the SSE is skipped
for any survivor with $K_s+1>K_t^{\rm best}$. Its validity must still be
checked because it remains relevant to the future candidate set. A later
candidate with a smaller $K_s$ is still evaluated, so no monotonicity of
$(K_s)$ is assumed.

### `subtests = "both"`: stop at the newest failure

If $\mathcal F_t\ne\varnothing$, let $m_t=\max\mathcal F_t$. Inclusive-left
pruning and right pruning together give

\begin{equation}
\label{eq:both-update}
\mathcal A_t=
\{s\in\mathcal A_{t-1}:s>m_t\}\cup\{t\};
\end{equation}

if there is no failure, all old boundaries survive. The optimized code scans
from the newest boundary towards zero. At the first invalid boundary it stops:
Equation~\eqref{eq:both-update} deletes that candidate and every untested older
candidate, so none can affect either the current optimum or a future candidate
set. The surviving suffix is then processed in increasing boundary order.
The code first finds its smallest $K_s+1$, then computes SSE only for candidates
in that segment-count class. The increasing pass preserves the previous
smallest-boundary rule under an exact numerical tie.

The reverse stopping rule is an implementation shortcut, not an additional
statistical pruning assumption. The statistical equality between `"both"`
and `"none"` still relies on Equations~\eqref{eq:right}--\eqref{eq:left}.

## Operation counts and consequences

Let $A_t=|\mathcal A_{t-1}|$, the stored count reported by `nb[t]`, and let
$E_t$ be the number of candidate validity states actually checked. The
candidate-loop work is controlled by $\sum_t E_t$, not necessarily by
`sum(nb)` $=\sum_t A_t$.

For `"none"`, $A_t=t$ and the stored exposure is exactly $n(n+1)/2$. If
$k_t^\star$ is defined above, however,

\[
E_t=\sum_{k<k_t^\star}|\mathcal B_{t-1,k}|
     +|\mathcal B_{t-1,k_t^\star}|.
\]

For `"right"`, the one-segment phase usually has $E_t=1$ even though the
physical vector grows; ending that phase incurs a one-off replay. Afterwards
$E_t=A_t$, but SSE is computed only when the candidate can match or improve
the provisional segment count. For `"both"`, if a failure occurs then

\[
E_t=|\{s\in\mathcal A_{t-1}:s\geq m_t\}|,
\]

whereas $E_t=A_t$ when every candidate is valid. Hence `"both"` benefits most
when a recent boundary fails; its reverse scan gives no saving at an endpoint
with no failure.

All three modes retain an $O(n^2)$ worst-case number of constant-time
candidate checks. The shortcuts improve the realized count and constants, not
that worst-case order. Test-specific work must also be included: Gaussian
FOCuS maintains a functional state, while the exact AR(1) statistic can scan a
whole candidate segment when requested and therefore has a larger unpruned
worst case. Under the stability and regular-segmentation assumptions of the
SVP paper (version 2), the non-decreasing sequence $(K_t)$ and bounded spans of
successive segment-count classes can instead yield the paper's near-linear
regime. Those assumptions are not inferred merely from observing a small
`nb` vector.

## Simulated behavior

The examples use `ts_generator()` with Gaussian noise; no constant-valued
series is used. First, consider four strong mean segments.

```{r strong-example, echo=TRUE}
set.seed(20260914)
y_strong <- ts_generator(
  chpts = c(50, 100, 150, 200),
  parameters = c(0, 3, -2, 2),
  sd_noise = 1, type = "gauss"
)
gamma_strong <- 2 * log(length(y_strong))
strong_fits <- setNames(lapply(modes, function(mode) {
  SVP(y_strong, gamma_strong, test = "gaussian_mean", subtests = mode)
}), modes)
```

```{r strong-table}
strong_table <- data.frame(
  Mode = modes,
  `segment endpoints` = vapply(strong_fits, endpoint_text, character(1)),
  K = vapply(strong_fits, function(z) length(z$changepoints), integer(1)),
  `sum(nb)` = vapply(strong_fits, function(z) sum(z$nb), numeric(1)),
  `final candidates` = vapply(
    strong_fits, function(z) length(z$lastIndexSet), integer(1)
  ),
  check.names = FALSE
)
knitr::kable(strong_table, format = "latex", booktabs = TRUE,
             row.names = FALSE)
```

```{r candidate-plot}
mode_colours <- c(none = "#000000", right = "#0072B2", both = "#D55E00")
mode_lty <- c(none = 1, right = 2, both = 3)
plot(strong_fits$none$nb, type = "l", log = "y", lwd = 1.5,
     col = mode_colours["none"], lty = mode_lty["none"],
     xlab = "Endpoint t", ylab = "Stored candidates nb[t]")
for (mode in modes[-1L]) {
  lines(strong_fits[[mode]]$nb, col = mode_colours[mode],
        lty = mode_lty[mode], lwd = 1.5)
}
legend("topleft", modes, col = mode_colours[modes], lty = mode_lty[modes],
       lwd = 1.5, bty = "n", cex = 0.82)
```

The following noisy no-change series shows why pruning assumptions matter.

```{r recovery-example, echo=TRUE}
set.seed(13001)
y_recovery <- ts_generator(
  chpts = 30, parameters = 0, sd_noise = 1, type = "gauss"
)
recovery_fits <- setNames(lapply(modes, function(mode) {
  SVP(y_recovery, gamma = 2, test = "gaussian_mean", subtests = mode)
}), modes)
```

```{r recovery-table}
recovery_table <- data.frame(
  Mode = modes,
  `segment endpoints` = vapply(recovery_fits, endpoint_text, character(1)),
  K = vapply(recovery_fits, function(z) length(z$changepoints), integer(1)),
  `sum(nb)` = vapply(recovery_fits, function(z) sum(z$nb), numeric(1)),
  check.names = FALSE
)
knitr::kable(recovery_table, format = "latex", booktabs = TRUE,
             row.names = FALSE)
```

Boundary zero is valid at $t=30$, so `"none"` returns one segment. It failed
at an earlier endpoint and was permanently deleted by `"right"`; `"both"`
removes still more boundaries. This is a witness for the endpoint-only rule,
not evidence that one statistical choice is universally preferable.

## Comparison with the existing `time_gaussian` simulation

The version 0.2.0 benchmark reruns the exact checked-in design: 20 logarithmic
sizes from 256 to 16,384 under no change (20 repetitions), and 0 to 100 changes
at $n=10{,}000$ (10 repetitions), for 5,640 paired fits. Every series is made
with `ts_generator()` using the same seeds and is shared by all methods. The
original `gaussian_time_*` files are preserved; new files use the
`gaussian_time_shortcuts_*` prefix. Pairing uses experiment, $n$, $k$,
repetition, and method.

```{r benchmark-check}
stopifnot(
  nrow(time_comparison) == 5640L,
  all(time_comparison$detected_equal[
    time_comparison$method != "SVP BIC multiscale"
  ])
)

largest_n <- max(time_summary$n[time_summary$experiment == "vary_n"])
largest_table <- subset(
  time_summary, experiment == "vary_n" & n == largest_n,
  select = c("method", "baseline_median", "shortcut_median", "median_speedup")
)
names(largest_table) <- c(
  "Method", "Baseline (s)", "Version 0.2.0 (s)", "Speedup"
)
largest_table[["Speedup"]] <- ifelse(
  largest_table[["Speedup"]] < 0.01,
  formatC(largest_table[["Speedup"]], format = "e", digits = 2),
  formatC(largest_table[["Speedup"]], format = "f", digits = 2)
)
knitr::kable(
  largest_table, format = "latex", booktabs = TRUE, digits = 3,
  caption = paste("Median time at n =", format(largest_n, big.mark = ","),
                  "with no change."),
  row.names = FALSE, position = "H"
)
```

The BIC and calibrated BIC methods use `"right"`; the multiscale method uses
`"both"`. PELT is an unchanged timing control. The segment counts agree in
`r sum(time_comparison$detected_equal)` of 5,640 rows. Every PELT and right-mode
row agrees. The `r sum(!time_comparison$detected_equal)` disagreements are all
multiscale no-change rows: the historical file was generated before the
current `subtests` API and subsequent multiscale candidate corrections. A
frozen build from immediately before the shortcut rewrite gives the same
output as version 0.2.0 on a mismatching historical series, so these 74 changes
must not be attributed to the new reverse scan.

The table shows that the two right-mode timings at the largest $n$ are
essentially unchanged. Thus this benchmark does not confirm a material speedup
from skipping their relatively cheap SSE calculations; validity-state work
dominates there. The old and new multiscale times are not an isolated shortcut
comparison because their algorithmic outputs differ, and the reverse scan has
no saving at an endpoint where every candidate is valid. Ratios are reported
only when both medians are positive, since the timer records some short fits as
zero. These are empirical wall-clock measurements, not complexity proofs.
The complete before/after and speedup curves are saved in
`simulations/time_gaussian/plots`.

\vspace{0.3em}
\textbf{Render environment.} `r R.version.string`; `svpChange2`
`r as.character(utils::packageVersion("svpChange2"))`;
`r R.version$platform`.

\textbf{Source basis.} Mathematical notation follows the SVP paper, version 2.
Implementation statements were checked against `src/svp.cpp` and
`src/svp_internal.h` in `svpChange2` 0.2.0 on 14 September 2026.
