---
title: "Corrected Section 5.3 AR(1) power study"
author: "svpChange2"
date: "2026-09-15"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
project_root <- normalizePath(
  file.path(dirname(knitr::current_input(dir = TRUE)), "..", "..")
)
knitr::opts_knit$set(root.dir = project_root)
setwd(project_root)
source(file.path("simulations", "power_ar1", "run_power.R"))
```

# Study model

The experiment follows the paper's Section 5.3 design:

\[
Y_t = \mu_t + E_t, \qquad E_t = \rho E_{t-1} + \eta_t,
\qquad \rho=0.8.
\]

The noise is generated by `ts_generator(type = "gaussAR1")` with
`sd_noise = sqrt(1-rho^2)`. Its stationary marginal variance is therefore one,
while the innovation variance supplied to DeCAFS and AR1Focus is
`sigma2 = 1-rho^2 = 0.36`. The first residual is drawn from the stationary
distribution.

# Signal design and metrics

The four signals are `none`, `up`, `updown`, and `rand1`, using the shared
`generate_power_signal()` implementation. We use `n = 600`, 100 replications,
and jump sizes `seq(1, 5, 0.2)`, giving 8,400 simulated series.

The mandatory terminal endpoint `n` is removed before scoring. Internal
changepoints are matched one-to-one within the Section 5.3 tolerance of five
observations. The study records precision, recall, F1, probability of the
correct number of changepoints, signal MSE, fitted segment count, and maximum
localization error when the number of changes is correct.

```{r design-check}
stopifnot(
  identical(AR1_N, 600L),
  identical(AR1_JUMP_SIZES, seq(1, 5, 0.2)),
  length(AR1_JUMP_SIZES) == 21L,
  AR1_REPS == 100L,
  AR1_TOLERANCE == 5L
)
```

# Calibration

The DeCAFS AR(1) penalty is `beta = c_D log(n)` with
`modelParam = list(sdEta = 0, sdNu = sqrt(1-rho^2), phi = rho)`. The smallest
candidate whose corrected null F1 is at least 0.99 is selected. The
approximate AR(1) PELT penalty and AR1Focus threshold are then selected to have
null F1 closest to the selected DeCAFS target. Calibration uses 1,000 null
series and validation uses 1,000 independent null series; both stages use
deterministic seeds and Wilson confidence intervals.

The approximate PELT candidate applies the existing PELT implementation to
the innovations

\[
Z_t = Y_{t+1} - \rho Y_t,
\]

with penalty `c_P log(n)`. A transformed endpoint `j` is mapped back to the
original endpoint `j+1`; the transformed terminal endpoint is discarded and
the original endpoint `n` is appended. This is approximate because it does not
retain the complete boundary correction of the exact AR(1) likelihood.

`PELT inflated` remains fixed at the paper-study value
`3*log(n)*(1+rho)/(1-rho)`. Its null F1 is reported as a fixed-baseline
benchmark and is not used as a calibration target. The approximate PELT
penalty is calibrated alongside DeCAFS and AR1Focus.

```{r calibration-summary, eval=FALSE}
calibration <- run_ar1_calibration(workers = 1L)
calibration$selected
calibration$validation
```

# Methods in the power study

The final four methods are exactly:

1. `PELT AR1 approximate`;
2. `PELT inflated`;
3. `DeCAFS AR1`;
4. `SVP AR1Focus / right / c = ...`.

The SVP call uses `test = "AR1Focus"` and `subtests = "right"` only. No
multiscale or `both` configuration is part of this study.

# Reproduction

Run from the package root:

```{r reproduction, eval=FALSE}
calibration <- run_ar1_calibration(workers = 8L)
run_and_save_ar1(workers = 8L, calibration = calibration)
source("simulations/power_ar1/export_paper_figures.R")
export_ar1_paper_figures(calibration = calibration)
```

The seven study PDFs are written to `simulations/power_ar1/plots/`. The
paper-facing Section 5.3 PDFs are written to the sibling `SVP_NEW_Figures2`
directory using the `5_` filename convention. The calibration table records
all candidates, null F1 values, confidence intervals, seeds, selected values,
and validation results.
