---
title: "AR(1) SVP models in `svpChange2`"
author: ""
date: "14 September 2026"
output:
  pdf_document:
    latex_engine: pdflatex
fontsize: 10pt
geometry: [a4paper, margin=2cm]
header-includes:
  - \usepackage{amsmath,amssymb,booktabs}
  - \DeclareMathOperator*{\arglexmin}{arg\,lex\,min}
  - \DeclareMathOperator*{\lexmin}{lex\,min}
  - \setlength{\parindent}{0pt}
  - \setlength{\parskip}{3pt}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.width = 6.7, fig.height = 2.35,
  out.width = "96%"
)
library(svpChange2)
```

This note describes every AR(1) interface used by SVP in `svpChange2`:
the generator, robust estimation of $\rho$, the exact fixed-variance and
profiled tests, the functional-pruning implementation, and their common SVP
recursion. Segment endpoints returned by `SVP()` are one-based and inclusive;
the last endpoint is always $n$.

## Data model and simulation

Let $\mu_t$ be piecewise constant and

\[
Y_t=\mu_t+E_t,\qquad E_t=\rho E_{t-1}+\eta_t,\qquad
\eta_t\stackrel{\mathrm{iid}}{\sim}\mathcal N(0,\sigma^2),\qquad |\rho|<1.
\]

Equivalently,
$Y_t=\mu_t+\rho(Y_{t-1}-\mu_{t-1})+\eta_t$.  The call
`ts_generator(type = "gaussAR1", rho = rho, sd_noise = sigma)` implements
this equation and initializes
$Y_1=\mu_1+\eta_1/\sqrt{1-\rho^2}$.  Thus `sd_noise` is the **innovation**
standard deviation and the matching SVP argument is
`sigma2 = sd_noise^2`; the stationary marginal variance is
$\sigma^2/(1-\rho^2)$.

If `rho = NA_real_`, the package computes once, from the complete series,

\[
M_k=\operatorname{median}_{t=k+1,\ldots,n}|Y_t-Y_{t-k}|,\qquad
\widehat\rho=\operatorname{clip}_{[-0.999,0.999]}
\left(\frac{M_2^2}{M_1^2}-1\right).
\]

For stationary Gaussian AR(1) noise,
$\operatorname{Var}(Y_{t+k}-Y_t)\propto1-\rho^k$, so
$M_2^2/M_1^2=1+\rho$. Medians make `AR1_rho()` resistant to the relatively
few differences crossing mean changes, although short or change-dense series
can still give an imprecise estimate.

## Exact conditional scan: `AR1`

Consider one candidate segment $x_1,\ldots,x_m$ and condition on $x_1$.  Put
$z_i=x_{i+1}-\rho x_i$, $i=1,\ldots,m-1$.  Under no change,
$z_i=(1-\rho)\mu+\eta_{i+1}$ and

\[
RSS_0=\min_\mu\sum_{i=1}^{m-1}\{z_i-(1-\rho)\mu\}^2
=\sum_{i=1}^{m-1}(z_i-\bar z)^2.
\]

For a change after observation $\tau$, $2\leq\tau\leq m-2$, write
$\beta=(\mu_1,\mu_2)^\top$ and define the $(m-1)\times2$ design

\[
(X_\tau)_{i\cdot}=
\begin{cases}
(1-\rho,0),&i<\tau,\\
(-\rho,1),&i=\tau,\\
(0,1-\rho),&i>\tau.
\end{cases}
\]

The middle row is essential: its residual is
$x_{\tau+1}-\rho x_\tau-\mu_2+\rho\mu_1$. Therefore

\[
RSS_1(\tau)=\min_\beta\|z-X_\tau\beta\|^2,\qquad
RSS_1=\min_{2\leq\tau\leq m-2}RSS_1(\tau),\qquad
T_{\mathrm{fix}}=\frac{RSS_0-RSS_1}{2\sigma^2}.
\]

`AR1_single_change()` returns these two RSS values, the minimizing $\tau$,
$T_{\mathrm{fix}}$, and `valid = (T_fix < gamma)`. Prefix sums of $z_i$ and
$z_i^2$ make each $RSS_1(\tau)$ calculation $O(1)$ and the complete
single-segment scan $O(m)$. Segments with fewer than four observations have
statistic zero. `valid_AR1()` evaluates the same equations in R and returns
only the validity indicator, for example for use as a test in `svp0()`.

\newpage

## Profiled scale and functional pruning

`test = "AR1Profile"` uses the same designs but maximizes separately over the
innovation variance under the null and alternative:

\[
T_{\mathrm{prof}}=\frac{m-1}{2}\log\left(\frac{RSS_0}{RSS_1}\right).
\]

It is exactly equivalent to `test = "AR1", profile_sigma = TRUE`. Because the
innovation variance is profiled out, `sigma2` is ignored in these two forms and
does not need to be positive.
Because $T_{\mathrm{prof}}$ and $T_{\mathrm{fix}}$ have different null laws,
their values of `gamma` require separate calibration.

`test = "AR1Focus"` uses fixed $\rho$ and $\sigma^2$ but avoids explicitly
rescanning every $\tau$. For each change location, the conditional
log-likelihood is a concave quadratic in $(\mu_1,\mu_2)$. The C++ recurrence
updates four functional envelopes (left/right and the two signs), retains only
quadratic pieces that can be optimal, and includes the crossing-transition
term above. If $\Delta_{\mathrm F}(\tau)$ denotes the quadratic gain retained
by this recurrence, its statistic is on the same half-likelihood-ratio scale:

\[
T_{\mathrm{Focus}}
=\frac{1}{2\sigma^2}\max_\tau\Delta_{\mathrm F}(\tau).
\]

Although the internal class comment calls this recurrence exact, the
user-facing help correctly treats it as an approximation: current numerical
paths do not satisfy $T_{\mathrm{Focus}}=T_{\mathrm{fix}}$ for every segment.
Therefore `AR1Focus` and `AR1` often agree, but endpoint equality is not
guaranteed; this point is quantified on page 4.

## The common SVP problem

For a segment $I_{s,t}=\{s+1,\ldots,t\}$, let
$V_\gamma(s,t)$ mean that the selected AR(1) statistic is strictly below
$\gamma$. All three choices solve

\[
\widehat\tau\in
\arglexmin_{\substack{0=\tau_0<\cdots<\tau_K=n\\
V_\gamma(\tau_{k-1},\tau_k)\ \forall k}}
\left(K,\sum_{k=1}^K C(\tau_{k-1},\tau_k)\right),
\quad
C(s,t)=\sum_{i=s+1}^{t}(Y_i-\bar Y_{s+1:t})^2.
\]

Thus the AR(1) likelihood determines **validity**, while ordinary
within-segment SSE is the secondary tie-breaker in the current generic SVP
implementation. With $R_t=(K_t,Q_t)$ and $R_0=(0,0)$,

\[
R_t=\lexmin_{0\leq s<t:\,V_\gamma(s,t)}
\{R_s+(1,C(s,t))\}.
\]

The exact test costs $O(m)$ whenever the statistic of a length-$m$ candidate
is requested. Consequently, without candidate pruning, the generic SVP loop
has $O(n^2)$ candidate--endpoint pairs but an $O(n^3)$ worst-case total for
`AR1` and `AR1Profile`. `AR1Focus` replaces the inner scan by functional
updates and is generally faster; its work depends on how many quadratic pieces
survive. `subtests = "right"` or `"both"` removes SVP candidates;
`"none"` is the assumption-free reference when the needed pruning
implications have not been established for a chosen statistic.

## Parameterization

| Purpose | Call and interpretation |
|:--|:--|
| fixed $\rho,\sigma^2$ | `SVP(y, gamma, "AR1", rho=rho, sigma2=sigma^2)` |
| unknown $\sigma^2$ | `SVP(y, gamma, "AR1Profile", rho=rho)` |
| same profiled test | `SVP(y, gamma, "AR1", rho=rho, profile_sigma=TRUE)` |
| functional recurrence | `SVP(y, gamma, "AR1Focus", rho=rho, sigma2=sigma^2)` |
| unknown $\rho$ | use `rho=NA_real_` or first compute `AR1_rho(y)` |
| R validity wrapper | `valid_AR1(y, gamma, rho, sigma2, profile_sigma)` |
| validity pruning | use `subtests="none"` for the unpruned benchmark |

\newpage

## One simulated series

The following reproducible example has $n=180$, changes after observations
60 and 120, $\rho=0.7$, $\sigma=0.8$, and $\gamma=6$. No constant-valued
segments are used: every observation contains Gaussian AR(1) noise.

```r
set.seed(20260914)
y <- ts_generator(c(60,120,180), c(0,2.5,-1.5),
                  sd_noise=.8, rho=.7, type="gaussAR1")
SVP(y, 6, test="AR1", subtests="none", rho=.7, sigma2=.8^2)
```

```{r example-data}
set.seed(20260914)
n <- 180L
true_ends <- c(60L, 120L, 180L)
true_mean <- rep(c(0, 2.5, -1.5), each = 60L)
y <- ts_generator(
  chpts = true_ends, parameters = c(0, 2.5, -1.5),
  sd_noise = 0.8, rho = 0.7, type = "gaussAR1"
)

configuration <- list(
  `AR1` = list(test = "AR1", rho = 0.7),
  `AR1Profile` = list(test = "AR1Profile", rho = 0.7),
  `AR1Focus` = list(test = "AR1Focus", rho = 0.7),
  `AR1; rho estimated` = list(test = "AR1", rho = NA_real_)
)
fits <- lapply(configuration, function(z) {
  SVP(y, 6, test = z$test, subtests = "none",
      rho = z$rho, sigma2 = 0.8^2)
})

example_table <- data.frame(
  Method = names(fits),
  `rho used` = vapply(fits, function(f) f$rho, numeric(1)),
  `inclusive segment endpoints` = vapply(
    fits, function(f) paste(f$changepoints, collapse = ", "), character(1)
  ),
  check.names = FALSE
)
rownames(example_table) <- NULL
knitr::kable(example_table, format = "latex", booktabs = TRUE, digits = 3)
```

```{r example-plot}
step_fit <- function(values, ends) {
  starts <- c(1L, head(ends, -1L) + 1L)
  unlist(Map(function(a, b) rep(mean(values[a:b]), b - a + 1L),
             starts, ends), use.names = FALSE)
}
plot(y, type = "l", col = "grey45", lwd = 0.7,
     xlab = "Observation", ylab = expression(Y[t]))
lines(true_mean, type = "s", lwd = 2, lty = 2, col = "black")
lines(step_fit(y, fits[[1]]$changepoints), type = "s",
      lwd = 2, col = "#0072B2")
abline(v = head(fits[[1]]$changepoints, -1L) + 0.5,
       col = "#0072B2", lty = 3)
legend("topleft", c("data", "true mean", "AR1 SVP fit"),
       col = c("grey45", "black", "#0072B2"),
       lty = c(1, 2, 1), lwd = c(0.7, 2, 2), bty = "n", cex = 0.82)
```

The exact, profiled, and functional implementations return the same
three-segment partition here; estimating $\rho$ gives
$\widehat\rho=`r sprintf("%.3f", fits[[4]]$rho)`$ and the same endpoints.

## Repeated simulation

To check behavior beyond one realization, the experiment was repeated 100
times with the same model and threshold. The table reports the proportion
with the correct three segments and the median Hausdorff distance $d_H$
between estimated and true internal changepoints. A fit with no internal
changepoint is penalized by $n$. These are finite-sample illustrations, not a
calibration of `gamma`; fixed and profiled statistics should be calibrated
separately for inferential comparisons.

```{r monte-carlo}
hausdorff_cp <- function(estimated, truth, n) {
  if (!length(estimated) || !length(truth)) return(n)
  max(
    vapply(estimated, function(x) min(abs(x - truth)), numeric(1)),
    vapply(truth, function(x) min(abs(x - estimated)), numeric(1))
  )
}

B <- 100L
method_names <- names(configuration)
correct_K <- matrix(FALSE, B, length(configuration))
distance <- matrix(NA_real_, B, length(configuration))
focus_equal <- logical(B)
rho_estimates <- numeric(B)
set.seed(20260914)

for (b in seq_len(B)) {
  yb <- ts_generator(
    chpts = true_ends, parameters = c(0, 2.5, -1.5),
    sd_noise = 0.8, rho = 0.7, type = "gaussAR1"
  )
  fb <- lapply(configuration, function(z) {
    SVP(yb, 6, test = z$test, subtests = "none",
        rho = z$rho, sigma2 = 0.8^2)
  })
  focus_equal[b] <- identical(fb[[1]]$changepoints, fb[[3]]$changepoints)
  rho_estimates[b] <- fb[[4]]$rho
  for (j in seq_along(fb)) {
    estimated <- head(fb[[j]]$changepoints, -1L)
    correct_K[b, j] <- length(estimated) == 2L
    distance[b, j] <- hausdorff_cp(estimated, c(60L, 120L), n)
  }
}

mc_table <- data.frame(
  Method = method_names,
  `Pr(correct K=3)` = colMeans(correct_K),
  `median d_H` = apply(distance, 2L, median),
  check.names = FALSE
)
knitr::kable(mc_table, format = "latex", booktabs = TRUE, digits = 2)
```

`AR1` and `AR1Focus` returned identical endpoint vectors in
`r sum(focus_equal)`/`r B` repetitions. For the runs estimating $\rho$,
$\operatorname{mean}(\widehat\rho)=`r sprintf("%.3f", mean(rho_estimates))`$
and $\operatorname{sd}(\widehat\rho)=`r sprintf("%.3f", sd(rho_estimates))`$.
The loss in correct segment count illustrates why a known or independently
estimated $\rho$ can matter, even though the median localization error remains
small in this example.

The simulations deliberately use `subtests = "none"`: they compare the three
statistics without confounding them with validity-pruning assumptions. For a
production analysis, calibrate `gamma` for the chosen statistic, variance
treatment, pruning mode, and sample size.

\newpage

## Which results should be identical?

The required parameter matching is summarized below. Here, identical means
the same returned endpoint vector, including $n$.

| Comparison | Expected result |
|:--|:--|
| `AR1Profile` versus `AR1` with `profile_sigma=TRUE` | Identical: both select the same C++ test class and profiled branch. |
| `rho=NA_real_` versus `rho=AR1_rho(y)` in the same test | Identical: the same full-series estimator is evaluated before SVP. |
| fixed-scale `AR1` versus `AR1Focus` | Often identical, but not guaranteed: they use different numerical recurrences. |
| fixed-scale `AR1` versus `AR1Profile` | Not expected: their statistics and threshold calibrations differ. |

The 100 moderate-signal simulations above happened to give identical `AR1`
and `AR1Focus` endpoints. A broader falsification check used
$n\in\{40,80,160\}$, $\rho\in\{-0.8,-0.4,0,0.3,0.7,0.9\}$,
$\gamma\in\{1.5,3,6,10\}$ and 20 seeds: 32 of 1,440 endpoint vectors differed.
This is numerical evidence that the equality is not universal, not a measure
of the methods' statistical quality.

```{r focus-falsification-code, include=FALSE, eval=FALSE}
# Reproducibility outline for the 1,440 comparisons.
for (rho in c(-0.8, -0.4, 0, 0.3, 0.7, 0.9))
  for (n in c(40L, 80L, 160L))
    for (seed in seq_len(20L)) {
      set.seed(100000 + seed + 10 * n + round(100 * rho))
      yy <- ts_generator(c(n %/% 3L, 2L * n %/% 3L, n),
                         c(-0.5, 1.7, -1.2), 0.9, rho,
                         type = "gaussAR1")
      for (g in c(1.5, 3, 6, 10)) {
        exact <- SVP(yy, g, "AR1", "none", 0.9^2, rho)
        focus <- SVP(yy, g, "AR1Focus", "none", 0.9^2, rho)
        # Record identical(exact$changepoints, focus$changepoints).
      }
    }
```

## Time complexity and empirical confirmation

For `AR1` and `AR1Profile`, testing a length-$m$ segment scans $O(m)$ change
locations. With no SVP pruning, the total work is therefore

\[
\sum_{t=1}^n\sum_{s=0}^{t-1}O(t-s)=O(n^3).
\]

For `AR1Focus`, let $P_{s,t}$ be the number of quadratic pieces inspected when
$I_{s,t}$ is extended. Its work is
$O(\sum_{s<t}P_{s,t})$: the worst case remains $O(n^3)$ because
$P_{s,t}\leq t-s$, but effective functional pruning typically keeps
$P_{s,t}$ small and gives near-$O(n^2)$ behavior.

The following median elapsed times (seconds) use five runs, R 4.6.1 on arm64,
the same noisy three-segment AR(1) model as above, and `subtests="none"` so all
$n(n+1)/2$ candidate--endpoint pairs are identical across methods.

```{r timing-table}
timing_table <- data.frame(
  n = c(500L, 1000L, 2000L),
  AR1 = c(0.033, 0.272, 2.181),
  AR1Profile = c(0.034, 0.274, 2.159),
  AR1Focus = c(0.013, 0.056, 0.253)
)
knitr::kable(timing_table, format = "latex", booktabs = TRUE, digits = 3)
timing_slopes <- vapply(timing_table[-1L], function(z) {
  unname(coef(lm(log(z) ~ log(timing_table$n)))[2L])
}, numeric(1))
```

The log--log slopes are `r paste(sprintf("%s %.2f", names(timing_slopes),
timing_slopes), collapse=", ")`. At $n=2000$, `AR1Focus` is
`r sprintf("%.1f", timing_table$AR1[3L] / timing_table$AR1Focus[3L])` times
faster than `AR1`. Thus the expected ranking is confirmed for a controlled,
large unpruned workload. With `right` or `both`, wall time also depends
on how many SVP candidates each statistic removes; timings are then not a pure
comparison of the three segment-test algorithms.

```{r timing-code, include=FALSE, eval=FALSE}
# Each n uses one generated data set; method/run order is randomized.
sizes <- c(500L, 1000L, 2000L)
methods <- c("AR1", "AR1Profile", "AR1Focus")
set.seed(20260914)
timing_data <- lapply(sizes, function(n) {
  ts_generator(c(n %/% 3L, 2L * n %/% 3L, n), c(0, 2.5, -1.5),
               sd_noise = 0.8, rho = 0.7, type = "gaussAR1")
})
design <- expand.grid(n = sizes, method = methods, run = seq_len(5L))
set.seed(314159)
for (i in sample(seq_len(nrow(design)))) {
  y_i <- timing_data[[match(design$n[i], sizes)]]
  gc(FALSE)
  tic <- proc.time()[[3L]]
  invisible(SVP(y_i, 6, design$method[i], "none", 0.8^2, 0.7))
  design$elapsed[i] <- proc.time()[[3L]] - tic
}
aggregate(elapsed ~ n + method, design, median)
```

\vspace{0.4em}
\textbf{Reference.} Chakar, S., Lebarbier, E., L\'evy-Leduc, C. and Robin, S.
(2017). A robust approach for estimating change-points in the mean of an
AR(1) process. \emph{Bernoulli}, 23(2), 1408--1447.
