---
title: "Review of the corrected AR(1) Section 5.3 study"
author: "SVP simulation audit"
date: "2026-09-15"
output:
  pdf_document:
    keep_tex: true
geometry: margin=0.8in
fontsize: 10pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
root <- normalizePath(file.path("..", ".."), mustWork = TRUE)
setwd(root)
source(file.path("simulations", "power_ar1", "run_power.R"))
calibration <- readRDS(file.path(
  "simulations", "power_ar1", "ar1_calibration.rds"
))
results <- readRDS(file.path("simulations", "power_ar1", "results.rds"))
legacy_root <- file.path(
  root, "simulations", "other_simus", "power_ar1_legacy_20260915"
)
old <- if (file.exists(file.path(legacy_root, "results.rds"))) {
  readRDS(file.path(legacy_root, "results.rds"))
} else NULL
```

# Study summary

The corrected study uses stationary Gaussian AR(1) noise with `n = 600`,
`rho = 0.8`, unit marginal variance, four patterns, jumps from 1 to 5 in
steps of 0.2, 100 replications, and a five-observation matching tolerance.
The final result contains `r nrow(results)` method rows and the four methods
`r paste(ar1_method_labels(calibration), collapse = ", ")`.

The mandatory endpoint `n` is removed before one-to-one metric matching. The
reported metrics are precision, recall, F1, correct-number probability, signal
MSE, fitted segment count, and localization error.

# Calibration

The selected DeCAFS multiplier is `r calibration$selected$decafs_constant` and
its calibration null F1 is `r round(calibration$selected$target_f1, 4)`. The
selected approximate-Pelt multiplier is
`r calibration$selected$approximate_constant`; the selected AR1Focus
multiplier is `r calibration$selected$svp_constant`.

DeCAFS is the calibration target at corrected null F1 0.99. The approximate
Pelt and AR1Focus candidates are selected by closest calibration null F1 to
that DeCAFS target. The inflated PELT method remains a fixed paper baseline.
Calibration and validation use 1,000 independent null series each, with
deterministic seeds and Wilson intervals.

```{r calibration-table}
knitr::kable(
  calibration$validation[, c("method", "candidate", "null_f1",
                             "null_f1_lower", "null_f1_upper")],
  digits = 4,
  caption = "Validation null F1 for the selected methods."
)
```

# PELT approximation and AR1Focus

The approximate AR(1) PELT method applies the package PELT implementation to
`Z_t = Y_{t+1} - rho Y_t`, then maps an endpoint `j` in the transformed series
to endpoint `j+1` in the original series. This is an approximation because the
complete AR(1) likelihood includes a boundary correction that is not retained
by ordinary transformed-data least squares.

AR1Focus is called with `test = "AR1Focus"` and `subtests = "right"`. No
`both` or multiscale SVP call is included in the study.

# Legacy comparison

```{r legacy, results='asis'}
if (!is.null(old)) {
  common_methods <- c(
    "PELT AR1 approximate", "PELT inflated", "DeCAFS AR1", "SVP AR1Focus"
  )
  normalise_method <- function(x) {
    x <- as.character(x)
    ifelse(grepl("^SVP AR1Focus", x), "SVP AR1Focus", x)
  }
  new_summary <- results |>
    dplyr::mutate(algorithm = normalise_method(algorithm)) |>
    dplyr::filter(algorithm %in% common_methods) |>
    dplyr::group_by(algorithm) |>
    dplyr::summarise(F1 = mean(F1), Precision = mean(Precision),
                     Recall = mean(Recall), CorrectNumCP = mean(CorrectNumCP),
                     MSE = mean(MSE), NumSegments = mean(NumSegments),
                     .groups = "drop")
  old_summary <- old |>
    dplyr::mutate(algorithm = normalise_method(algorithm)) |>
    dplyr::filter(algorithm %in% common_methods) |>
    dplyr::group_by(algorithm) |>
    dplyr::summarise(F1 = mean(F1), Precision = mean(Precision),
                     Recall = mean(Recall), CorrectNumCP = mean(CorrectNumCP),
                     MSE = mean(MSE), NumSegments = mean(NumSegments),
                     .groups = "drop")
  print(knitr::kable(new_summary, digits = 4,
                     caption = "Corrected AR(1) study summary."))
  print(knitr::kable(old_summary, digits = 4,
                     caption = "Archived pre-refresh AR(1) study summary."))
}
```

# Reproduction

```r
source("simulations/power_ar1/calibrate_ar1.R")
calibration <- run_ar1_calibration(workers = 8L)
source("simulations/power_ar1/run_power.R")
run_and_save_ar1(workers = 8L, calibration = calibration)
source("simulations/power_ar1/export_paper_figures.R")
export_ar1_paper_figures(calibration = calibration)
```
