---
title: "Gaussian SVP: implementation and simulation review"
author: ""
date: "15 September 2026"
output:
  pdf_document:
    latex_engine: pdflatex
fontsize: 9pt
geometry: [a4paper, margin=1.8cm]
header-includes:
  - \usepackage{amsmath,amssymb,booktabs,array,float,longtable}
  - \DeclareMathOperator*{\arglexmin}{arg\,lex\,min}
  - \DeclareMathOperator*{\lexmin}{lex\,min}
  - \setlength{\parindent}{0pt}
  - \setlength{\parskip}{2.5pt}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.pos = "H", out.width = "98%"
)
library(svpChange2)
library(dplyr)
library(tidyr)

project_root <- normalizePath(file.path("..", ".."))
knitr::opts_knit$set(root.dir = project_root)
calibration <- read.csv(file.path(
  project_root, "simulations", "power_gaussian",
  "gaussian_mode_calibration.csv"
))
power <- readRDS(file.path(
  project_root, "simulations", "power_gaussian", "results.rds"
))
decomposition <- read.csv(file.path(
  project_root, "simulations", "power_gaussian",
  "gaussian_performance_decomposition.csv"
))
changes <- read.csv(file.path(
  project_root, "simulations", "power_gaussian",
  "gaussian_performance_changes.csv"
))
slopes <- read.csv(file.path(
  project_root, "simulations", "time_gaussian", "gaussian_time_slopes.csv"
))
time_summary <- read.csv(file.path(
  project_root, "simulations", "time_gaussian",
  "gaussian_time_summary_by_n.csv"
))
figure_root <- normalizePath(
  file.path(project_root, "..", "SVP_NEW_Figures2"), mustWork = FALSE
)
fmt <- function(x, digits = 3) formatC(x, digits = digits, format = "f")
```

## Scope and mathematical target

For independent observations $Y_i\sim\mathcal N(\mu_i,\sigma^2)$, write
$y_{s..t}=(y_{s+1},\ldots,y_t)$ and

\[
 \mathcal C(s,t)=\sum_{i=s+1}^{t}(y_i-\bar y_{s..t})^2,
 \qquad
 f(s,t)=\frac{1}{2\sigma^2}\max_{s<u<t}
 \{\mathcal C(s,t)-\mathcal C(s,u)-\mathcal C(u,t)\}.
\]

`SVP(..., test="gaussian_mean")` uses $\sigma^2=1$ and the half
log-likelihood-ratio scale above.  The argument `gamma` is a **validity
threshold**, not the additive penalty in PELT.  In the simulations it is
$\gamma=c\log n$, where $c$ is the free calibration constant printed in every
SVP legend.

For boundaries $0=\tau_0<\cdots<\tau_K=n$, define the mode-dependent segment
statistic

\[
F_0(s,t)=f(s,t),\quad
F_R(s,t)=\max_{s<u\le t}f(s,u),\quad
F_B(s,t)=\max_{s\le a<b\le t}f(a,b).
\]

Here $F_0,F_R,F_B$ correspond respectively to `none`, `right`, and `both`.
Singleton segments are accepted.  With strict validity $F_M(s,t)<\gamma$,
each mode solves

\[
\widehat\tau_M\in
\arglexmin_{\substack{0=\tau_0<\cdots<\tau_K=n\\
F_M(\tau_{j-1},\tau_j)<\gamma,\ j=1,\ldots,K}}
\left(K,\sum_{j=1}^{K}\mathcal C(\tau_{j-1},\tau_j)\right).
\]

Thus the modes are three related statistical procedures, not three generally
interchangeable implementations.  For a fixed $\gamma$,
$F_0\le F_R\le F_B$, so the feasible sets are nested and the optimal segment
counts satisfy $K_0\le K_R\le K_B$.  Equality of returned endpoints requires
the same validity definition and tie rule; calibration to the same null error
rate does not make the segmentations identical.

With $R_t=(K_t,Q_t)$ and $R_0=(0,0)$, the dynamic program is

\[
R_t=\lexmin_{\substack{0\le s<t\\F_M(s,t)<\gamma}}
\{R_s+(1,\mathcal C(s,t))\}.
\]

The compiled code compares SSE strictly and scans surviving boundaries in
increasing order, so an exact numerical tie keeps the earliest visited
boundary.

## What the Gaussian implementation does

Each candidate boundary $s$ owns an incremental test state, the last endpoint
already supplied to that state, and a validity flag.  Arrays `K`, `Q`, and
`previous` store $R_t$ and allow backtracking from $n$.  Centered first- and
second-moment prefix sums evaluate $\mathcal C(s,t)$ in constant time.
`GaussianFocusState` maintains two functional-pruning envelopes and returns
the maximum Gaussian gain over split points.  Its expected number of active
pieces is logarithmic under the usual independent continuous model, while its
worst case is linear in the candidate-segment length.

The three implementation shortcuts are exact for their stated validity rule:

* `none` keeps all boundaries, groups them by $K_s$, and tests groups in
  increasing $K_s$.  It stops after the first group containing a valid
  candidate, because larger groups cannot improve the primary coordinate.
* `right` returns immediately while boundary zero remains valid.  Later
  candidates are deferred; if zero fails, their missed prefix history is
  replayed exactly.  Afterwards invalid candidates are deleted and SSE is
  skipped whenever $K_s+1$ already exceeds the best segment count.
* `both` scans candidates from newest to oldest.  At the newest invalid
  candidate it stops and deletes that candidate and every older one.  It first
  finds the smallest surviving $K_s+1$, then computes SSE only in that class.

There is no whole-series preflight.  In particular, a no-change series does
not make `both` return before the dynamic-programming loop.

The other Gaussian-model entries in `SVP()` share the same lexicographic
engine but not the same validity statistic:

```{r gaussian-methods-table, results='asis'}
methods <- data.frame(
  test = c("gaussian_mean", "gaussian_variance", "AR1", "AR1Profile", "AR1Focus"),
  model = c(
    "independent mean changes, fixed variance 1",
    "zero-mean variance changes",
    "AR(1) mean changes, fixed rho and innovation variance",
    "exact AR(1), innovation variance profiled",
    "approximate AR(1) innovation transform"
  ),
  endpoint_check = c(
    "expected O(log m), worst O(m)",
    "O(a_m), worst O(m)",
    "O(m)", "O(m)", "O(a_m), worst O(m)"
  ), check.names = FALSE
)
knitr::kable(methods, booktabs = TRUE, longtable = FALSE,
             col.names = c("`test`", "model", "one candidate of length $m$"))
```

`AR1Focus` can differ from `AR1` even with known parameters because its
innovation split omits the exact boundary transition. `AR1Profile` can also
differ from fixed-variance `AR1`.  All variants currently use ordinary SSE as
the secondary tie-breaker, even when that is not their model likelihood.

## Correctness audit: changes recommended in the package

The Gaussian mean statistic is mathematically translation invariant, but its
FOCuS state currently accumulates the raw observations.  Centering is applied
only to the SSE prefix sums.  The following deterministic check shows the
problem (the rows report segment endpoints):

```{r translation-table, results='asis'}
set.seed(2718)
y <- rnorm(48)
gamma <- 2 * log(length(y))
translation <- do.call(rbind, lapply(c("none", "right", "both"), function(mode) {
  do.call(rbind, lapply(c(0, 1e6, 1e9, 1e12), function(offset) {
    data.frame(
      mode = mode, offset = format(offset, scientific = TRUE),
      endpoints = paste(SVP(y + offset, gamma, "gaussian_mean", mode)$changepoints,
                        collapse = ", ")
    )
  }))
}))
knitr::kable(translation, booktabs = TRUE)
```

The repair should center the observations **before they are passed to both the
validity state and the prefix-sum cost calculation**, preferably in the
Gaussian-mean dispatch path.  Add exact endpoint and predecessor tests for
positive and negative offsets $10^2,10^6,10^9,10^{12}$ in every mode.  The
existing $10^6$ check is not sufficiently adversarial.

Three further points require either a code change or an explicit paper
qualification:

1. `GaussianVariance` sends $Y_i^2$ to `GammaRate(shape=1)`.  If
   $Y_i\sim\mathcal N(0,\sigma^2)$, then $Y_i^2$ is Gamma with shape $1/2$.
   The constructor should use shape $1/2$, followed by null and power tests.
2. Validity is implemented as `statistic < gamma`; equality fails.  The paper
   must use the same strict inequality, or the implementation and boundary
   tests must be changed together.
3. Proposition 2 applies to the final-segment rule `none` on the stated cost
   scale.  It does not automatically justify a `right` or `both` curve, and a
   PELT SSE penalty and the half-likelihood SVP threshold must not be called
   the same parameter merely because both contain $2\log n$.

The public default is presently `both`.  Therefore, the paper's basic SVP
definition agrees with the software default only if it uses $F_B$; otherwise
the paper should request `subtests="none"` explicitly.  No package behavior was
changed in this review.

## Recalibration and Section 5.1

The calibration target is the probability of at least one internal false
change under $n=1000$ independent standard Gaussian observations.  PELT uses
its Manual penalty $2\log n$.  For each SVP mode, a grid search chooses $c$ in
$\gamma=c\log n$ to match PELT on 20,000 calibration series; 10,000 new series
are used only for validation.

```{r calibration-table, results='asis'}
cal_table <- calibration |>
  transmute(
    mode, c = constant,
    `calibration FPR` = svp_fpr_calibration,
    `validation FPR` = svp_fpr_validation,
    `PELT validation FPR` = pelt_fpr_validation,
    `absolute validation gap` = absolute_gap_validation
  )
knitr::kable(cal_table, digits = 4, booktabs = TRUE)
```

The validation Wilson intervals overlap for all three modes.  Section 5.1
keeps four curves: PELT, `SVP Gaussian / right / c = 2.000`, calibrated
`right`, and calibrated `both`.  Calibrated `none` is used in the runtime study
only.  Every series is generated once and shared across methods.

The scoring code now removes the mandatory endpoint $n$ before one-to-one
tolerance matching.  The earlier code treated $n$ as a true and detected
change, inflating F1.  For example, the mean PELT F1 over the jump grid changes
from 0.658 to 0.591 in `rand1`, although the fitted endpoints are identical.
PELT and `right, c=2` reproduce exactly after rescoring.  Recalibration changes
the two calibrated fits; its largest F1 change after holding the scoring rule
fixed is small in a change scenario.

```{r result-table, results='asis'}
stage_table <- decomposition |>
  filter(stage != "historical stored score") |>
  select(stage, configuration, pattern, mean_F1) |>
  pivot_wider(names_from = pattern, values_from = mean_F1)
knitr::kable(stage_table, digits = 3, booktabs = TRUE)
```

```{r f1-figure, out.width='96%', fig.cap='Refreshed Section 5.1 F1 curves. SVP labels report the test name, subtests mode, and free constant.'}
knitr::include_graphics(
  file.path(figure_root, "1_f1_vs_jump.pdf"), rel_path = FALSE
)
```

The revised conclusion is more cautious than the current paper.  PELT has the
best average F1 in the three change scenarios.  The calibrated SVP curves are
close to PELT in `rand1` around jump 0.9 and in `up`/`updown` near jump 2, but
they remain conservative at small and moderate jumps.  For `rand1` at jump 2,
mean F1 is about 0.87 for PELT and 0.84 for calibrated `both`, so the methods
are not uniformly indistinguishable above jump 0.9.

\newpage

## Figure 5: time and complexity

The refreshed benchmark uses 20 logarithmically spaced sizes from 256 to
16,384 with 20 replications under no change, and $n=10{,}000$ with 0--100 true
changes and 10 replications.  Data come from `ts_generator()`.  Methods receive
the same series in randomized order, after warm-up.  Calls shorter than 0.05 s
are batched and divided by the batch size.  Curves show median elapsed time;
bands are nonparametric bootstrap intervals.

```{r runtime-figure, out.width='100%', fig.cap='Controlled Gaussian runtime comparison. The exponents printed in panel (a) are descriptive log--log fits, not worst-case bounds.'}
knitr::include_graphics(
  file.path(figure_root, "Figure5_gaussian_runtime.pdf"), rel_path = FALSE
)
```

```{r slope-table, results='asis'}
knitr::kable(slopes, digits = 3, booktabs = TRUE,
             col.names = c("method", "range", "exponent", "2.5%", "97.5%"))
```

Let $E_t$ be the number of candidate validity states actually checked at
endpoint $t$.  Gaussian running time is described by
$\sum_t O(E_t\log t)$ in the usual expected FOCuS regime, with a larger worst
case when active functional pieces grow linearly.  The shortcuts determine
$E_t$:

* Under no change, `right` usually tests only boundary zero, hence expected
  $O(n\log n)$ validity work.  The historical sub-quadratic SVP curve was also
  a prefix-validity implementation with this one-segment behavior; its slope
  near 1.4 was empirical, affected by fixed overhead, coarse millisecond
  timings, and a short finite range.  It was not evidence for a universal
  $n^{1.4}$ complexity.
* `none` stops in the $K_s=0$ bucket only while boundary zero is valid.  Once
  that candidate fails, the $K_s=1$ bucket can contain many boundaries and all
  valid candidates in that bucket must be compared for the SSE tie-breaker.
  At the calibrated threshold these episodes produce a noisy, nearly
  quadratic curve (full-grid exponent about 1.92); retaining candidates lets
  boundary zero recover, but does not erase the cost already incurred.
* `both` has no zero-boundary shortcut.  If all recent candidates are valid,
  the reverse scan examines them all.  Thus the no-change candidate exposure
  is quadratic and the FOCuS factor gives expected $O(n^2\log n)$ in this
  regime.  It becomes advantageous only when a recent failure deletes a long
  prefix of candidates.

The old whole-series preflight had hidden this cost by returning the one-segment
answer before dynamic programming; it has been removed.  Figure 5 must
therefore not claim that `both` is uniformly the fastest or that its pruning
caps the candidate count under no change.  PELT's observed slope is likewise
implementation- and regime-specific; its standard pruning gives linear
expected time under assumptions, although adverse cases can be quadratic.

## Required edits to `SVP_NEW.pdf`

Section 5.1 should (i) call $\gamma$ a validity threshold, (ii) define
$\gamma=c\log n$, (iii) state the four plotted methods and calibrated constants,
(iv) describe the null-FPR calibration/validation samples, and (v) define F1
using internal changepoints and one-to-one matching.  Replace the claim of
near-equivalence above jump 0.9 by the result-specific comparison given above.

Section 5.4 should replace its three historical SVP configurations by
calibrated `none`, `right`, and `both`; state the controlled timing protocol;
and distinguish empirical exponents from bounds.  The explanation of `both`
must reflect the absent preflight and its quadratic no-change candidate scan.
The Figure 5 caption should identify medians, bootstrap intervals, hardware,
and the exact package versions.

Finally, revise the general method section so the base feasibility rule and
the software `subtests` value agree; qualify Proposition 2 as applying to the
`none` rule and the correct objective scale; use strict validity consistently;
and correct the statement that the `changepoint` PELT implementation is based
on Rcpp.  A standalone LaTeX replacement memo accompanies this document.
