---
title: "Review of the corrected heavy-tail Section 5.2 refresh"
author: "SVP simulation audit"
date: "2026-09-15"
output:
  pdf_document:
    keep_tex: true
geometry: margin=0.8in
fontsize: 10pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(dplyr)
root <- if (file.exists(file.path("simulations", "power_robust"))) {
  normalizePath(".", mustWork = TRUE)
} else {
  normalizePath(file.path("..", ".."), mustWork = TRUE)
}
setwd(root)
source_path <- if (file.exists(file.path(
  "simulations", "power_robust", "run_power.R"
))) file.path("simulations", "power_robust", "run_power.R") else
  file.path("..", "power_robust", "run_power.R")
source(source_path)
robust_root <- file.path(root, "simulations", "power_robust")
legacy_root <- file.path(
  root, "simulations", "other_simus", "power_robust_legacy_20260915"
)
calibration <- readRDS(file.path(robust_root, "robust_calibration.rds"))
calibration_table <- read.csv(
  file.path(robust_root, "true_true_calibration.csv"),
  stringsAsFactors = FALSE
)
new <- readRDS(file.path(robust_root, "results.rds"))
old <- readRDS(file.path(legacy_root, "results.rds"))
new <- new |> filter(algorithm %in% c(
  "PELT", "RFPOP", robust_method_labels(calibration)[[3L]],
  robust_method_labels(calibration)[[4L]]
))
old <- old |> filter(algorithm %in% c(
  "PELT", "RFPOP paper", "SVP MedianMood", "SVP Wilcoxon"
))
rename_old <- c(
  "PELT" = "PELT", "RFPOP paper" = "RFPOP",
  "SVP MedianMood" = "SVP MedianMood", "SVP Wilcoxon" = "SVP Wilcoxon"
)
rename_new <- c(
  "PELT" = "PELT", "RFPOP" = "RFPOP",
  setNames("SVP MedianMood", robust_method_labels(calibration)[[3L]]),
  setNames("SVP Wilcoxon", robust_method_labels(calibration)[[4L]])
)
old$method <- unname(rename_old[old$algorithm])
new$method <- unname(rename_new[new$algorithm])
metrics <- c("F1", "Precision", "Recall", "CorrectNumCP", "MSE", "NumSegments")
comparison <- full_join(
  new |> group_by(pattern, jump, method) |>
    summarise(across(all_of(metrics), mean), .groups = "drop") |>
    rename_with(~ paste0("new_", .x), all_of(metrics)),
  old |> group_by(pattern, jump, method) |>
    summarise(across(all_of(metrics), mean), .groups = "drop") |>
    rename_with(~ paste0("old_", .x), all_of(metrics)),
  by = c("pattern", "jump", "method")
) |>
  mutate(across(starts_with("new_"), as.numeric),
         across(starts_with("old_"), as.numeric)) |>
  mutate(delta_F1 = new_F1 - old_F1,
         delta_Precision = new_Precision - old_Precision,
         delta_Recall = new_Recall - old_Recall,
         delta_CorrectNumCP = new_CorrectNumCP - old_CorrectNumCP,
         delta_MSE = new_MSE - old_MSE,
         delta_NumSegments = new_NumSegments - old_NumSegments)
write.csv(comparison, file.path(root, "simulations", "z_robust",
                                 "paired_comparison.csv"), row.names = FALSE)
selected_comparison <- comparison |>
  group_by(pattern, method) |>
  summarise(across(starts_with("delta_"), mean), .groups = "drop") |>
  rename(
    dF1 = delta_F1, dPrecision = delta_Precision, dRecall = delta_Recall,
    dCorrect = delta_CorrectNumCP, dMSE = delta_MSE, dSegments = delta_NumSegments
  )
selected_table <- data.frame(
  method = c("RFPOP", "Wilcoxon", "MedianMood"),
  selected_parameter = c(
    calibration$selected$rfpop_constant,
    calibration$selected$wilcoxon_constant,
    calibration$selected$mood_alpha
  )
)
```

# Scope and acceptance checks

The regenerated study has `r nrow(new)` method rows, `r length(unique(new$algorithm))`
canonical algorithms, four patterns, 40 jump sizes, and 100 replications.
The calibration uses `r calibration$selected$calibration_reps` calibration and
`r calibration$selected$validation_reps` validation null series. The nominal
RFPOP target is `r calibration$target_nominal` corrected null F1; the selected
calibration estimate is `r round(calibration$selected$rfpop_target_f1, 5)`.

```{r selected}
knitr::kable(selected_table, digits = 6,
             caption = "Selected robust calibration parameters.")
```

The Gaussian-PELT external null-F1 benchmark is
`r round(calibration$gaussian_pelt_benchmark$gaussian_pelt_null_f1, 5)`.
The intentionally misspecified Gaussian PELT fit evaluated on the Student-t(2)
validation stream has null F1
`r round(calibration$student_pelt_benchmark$student_pelt_null_f1, 5)`.

# Calibration table

The full candidate table, including Wilson intervals, seeds, target, and
selection flags, is stored in `true_true_calibration.csv` in the robust study
folder. The
selected validation rows are shown here.

```{r calibration}
calibration_table |>
  filter(stage == "validation") |>
  select(method, candidate, null_f1, null_f1_lower, null_f1_upper,
         rfpop_target_f1, selected, seed) |>
  knitr::kable(digits = 5,
               caption = "Independent validation of selected null thresholds.")
```

# Paired before/after comparison

The detailed comparison is in `paired_comparison.csv`, indexed by pattern,
jump, and method. The table below reports the mean change over jump sizes for
each pattern. Positive values improve the metric; for MSE and segment count,
negative values are preferable.

```{r comparison}
selected_comparison |>
  knitr::kable(digits = 4,
               caption = "Mean new-minus-legacy metric change by pattern and method.")
```

# Revised Section 5.2 wording

Section 5.2 should describe the Student-t(2) experiment as a stress test with
the fixed Gaussian PELT baseline and calibrated robust procedures. All noise is
generated by `ts_generator(type = "student", df = 2, scale = 1)` for
`n=1000`, four signal patterns, jump sizes `0.1,...,4`, and 100 replications.
The endpoint `n` is excluded before one-to-one matching within the framework
tolerance. Precision, recall, F1, correct-number probability, MSE, segment
count, and localization error are then computed from the same metric code as
the Gaussian study.

RFPOP uses biweight loss (the `robseg` name is `loss = "Outlier"`) with
`lthreshold = 3` and `lambda = c_RF log(n)`. Its multiplier is calibrated on
independent t(2) null streams for at least 0.99 corrected null F1. The
Median-Mood and Wilcoxon SVP thresholds are independently selected to match
that RFPOP target and use `subtests = "right"` only. Their K-dependent formulas
use `oracle_segments`, the known data-generating number of segments in this
power experiment; this oracle is not available in ordinary applications.
